---
title: "Class 12 Pt2."
author: "Parisa Charoensook A17939591"
format: pdf
editor: visual
---

> Q13: Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes. 

The sample sizes for each genotype are 
A/A = 108, A/G = 233, and G/G = 121. The median expression levels were A/A = 31.25, A/G = 25.06, G/G = 20.07. 


```{r}
geno <- read.table("expression.txt" ,
header = TRUE) 
``` 

```{r}
table(geno$geno)

tapply(geno$exp, geno$geno, median)
```

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3? 

The boxplot shows that ORMDL3 expression decreases from A/A to A/G to G/G. Individuals with A/A genotype have highes expressions and G/G has least. This sugggests that the SNP is associated with changes in ORMDL3 expression and has an affect on gene expression. 

```{r}
boxplot(exp ~ geno, data=geno,
        main = "ORMDL3 Expression by Genotype",
        xlab = "Genotype",
        ylab = "Expression")
```


