---
title: "Class 5: Data Viz with ggplot2"
author: "Parisa Charoensook A17939591"
format: pdf
toc: true
---

## Background

There are many graphics systems in R for making plots and figures. These include so-called *"base R" graphics* like the `plot()` function and add on packages like **ggplot2**

Let's compare how we make a simple plot figure with these two systems: 

We can use the in-built `cars` dataset:

```{r}
head ( cars )
```
```{r}
plot(cars)
```

Before I can use ggplot2 I need to install it on my computer. 
To do this we can use the function `install.packages("ggplot2")`
> **N.B.** We never run `install.packages()` in our quarto doc (we run it once only in our R console) as it would re-install th epackage every time we render our quarto report. 

Once installed we need to load up the package into our R brain:
```{r}
library(ggplot2)
```


The main function in the **gglot2** package is called `ggplot()`

```{r}
ggplot(cars)
```

every ggplot has at least 3 layers: 

- the **data (a data.frame of the stuff we want to plot)
- the **aes**thetics (how the data maps to the plot)
- the **geom** layer (how you want the plot drawn)

```{r}
ggplot(cars) +
  aes(x=speed , y=dist ) +
  geom_point()
```

## Add some custom plot

Let's add a trend line that shows the relationship between speed and distance. 

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  labs(title="Speed and Stopping Distances of Cars",
       x="Speed (MPH)", 
       y="Stopping Distance (ft)") +
  geom_smooth(method="lm", se=FALSE) + theme_bw()
```

Q. Can you make the `geom_smooth()` function produce a linear straight line fit to the data and turn off the "gra" error region. 


---

##Gene expression figure 

Import the data to the plot
```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

```{r}
sum(genes$State == "up")
```

A useful new funciion in this contetx is the `table()` function:
```{r}
table(genes$State)
```

My first plot attempt
```{r}
ggplot(genes) + 
  aes(Condition1, Condition2, col=State)+
  geom_point() +
  scale_colour_manual(values=c("purple", "gray", "orange")) + theme_bw() +
  labs(x= "No drug", y="Drug", 
       title= "Expression changes upon GLP-1 inhibitor treatment") 
```

## Going further

Here we read the famous 
```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

>Q. How many entries (i.e. rows) are in this dataset?

```{r}
nrow(gapminder)

```

>Q. How many "country" entries are in this dataset?

```{r}
length( unique(gapminder$country))
```

Let's make our first plot of the entire dataset: 

Plot of "gdpPercap" vs: "lifeExp" colord by "continent"

```{r}
p <- ggplot(gapminder)+
  aes(gdpPercap, lifeExp, col = continent)+ 
  geom_point(alpha=0.3)
```
```{r}
p
```

I can add more layers to 'p' 
```{r}
p + 
  facet_wrap(~gapminder$continent)
```

Make a plot for years 1977 and 2007 only (not all the years in the dataset). 

>Q. First use the **dyplyr** package and the `filter()` function from that package to extract the ros from year 2007. 

```{r, message=FALSE, warning=FALSE}
library(dplyr)
```


```{r}
g07 <- filter(gapminder, year == 2007)
g77 <- filter(gapminder, year == 1977)
g <-filter(gapminder, year == 2007 | year == 1977)
```

```{r}
ggplot(g)+ 
  aes(gdpPercap, lifeExp, col=continent, size=pop) +
        geom_point() + facet_wrap(~year)
    
```

> Q. Make a histogram of lifeExp colored by continent (try using `fill=continent` or `col=continent`) 

```{r}
ggplot(gapminder)+ 
  aes(lifeExp, fill=continent) + geom_histogram()
```

```{r}
ggplot(gapminder)+ 
  aes(lifeExp, col=continent) + geom_histogram()
```


> Q. Make a histogram of life Exp faceted by continent

```{r}
ggplot(gapminder)+ aes(lifeExp)+ geom_histogram()+facet_wrap(~continent)
```




